---
title: "`epicontacts`: Visualisation of Network Data in Epidemiology"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
runtime: shiny
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Introduction

The `epicontacts` package aims to facilitate the visualisation and analysis of any epidemiological data with a network component, such as contact tracing datasets and transmission chains.

Input data must consist of two components:

<bk>

 1. A **line list** (LL) of individuals. The line list contains a first column with unique identifier of individuals (record of cases and/or contacts) and any attributes in additionnal columns (e.g. age, sex, location, etc.).

<bk>

2. A **contact dataset** (CD) mapping the relationships between individuals. The contact dataset corresponds to the mappings of the relationships between individuals (edges between nodes). The contact dataset can contain any attributes qualifying this relationship (e.g. exposure, serial interval, etc.).

<bk>

The package provides several network statistics, functions to subset networks (by edge or two nodes) and visualise networks in a various ways (2D, 3D, interactive graphics) using third-party packages.

---------------------------------------
  
### Installation
  
To install the development version from GitHub:
```{r install, eval=FALSE} 
library(devtools)
install_github("Hackout3/epicontacts")
```
  
The stable version can be installed from CRAN using: [*to be checked if needed*] _{package is not on CRAN yet, so prob not needed}_

```{r, eval=FALSE}
# install.packages("epicontacts",force = TRUE) 
```

To load the package use the following:

```{r, eval=TRUE} 
library(epicontacts)
```
  
### Overview

Main features include:

* **epi_contacts**: a new S3 class for storing line list and contacts datasets

* **as.igraph.epi_contacts**: creates an `igraph` object from a `epi_contacts` object

* **get_clusters**: assign clusters and corresponding cluster sizes to line list of an `epi_contacts` object (clusters being groups of connected individuals/nodes)

* **ebola.sim**: simulated Ebola outbreak dataset

* **get_degree**: number of contacts (edges)  per individuals (nodes) in a `epi_contacts` object

* **get_id**:access unique IDs identifier(s) from `epi_contacts` object with various options

* **get_pairwise**:  form an `epi_contacts` object, extract attributes of record(s) in contacts database using information provided in the line list database

* **graph3D**: 3D graph from an `epi_contacts` object

* **make_epi_contacts**: a constructor for the new `epi_contacts` class

* **subset_clusters_by_id**: subset an `epi_contacts` object based on individuals/nodes of interest

* **subset_clusters_by_size**: subset an `epi_contacts` object based on size(s) of clusters (clusters being groups of connected individuals/nodes)

* **vis_epi_contacts**: plots epi_contacts objects using the visNetwork package

* **epicontacts_server**: launches the epicontacts Shiny application on a local host

Generic functions include:

* **summary**: summary for  `epi_contacts` object

* **plot**:  default plot for `epi_contacts` object using visNetwork package

* **print**: default printing method for `epi_contacts` object

* **x[i,j,contacts]**: subset an `epi_contacts` object by retaining specified cases

### Package Contents

```{r Overv, eval=TRUE}
ls("package:epicontacts")
```

Load additional packages required for running examples below: 

```{r lib}
library(shiny)
```

### Visualisation and Analysis

##### Load training dataset
  
A new S3 class `epi_contacts` for storing line list (LL) and contacts dataset (CD) can be created from a line list (one row = one individual) and a contact dataset containing the relationship betwen individuals (one row = one edge). A training dataset `mers_kor_14` is used as running example (for more information, see` outbreaks` package  for dataset description).

```{r , eval=TRUE}
str(mers_kor_14) # list of two dataframes: 1) line list and 2)contact dataset.
```

Dataframe corresponding to a line listing of individuals (= network nodes) which can be case(s) and/or contact(s) (if any), unique identifier and a set of descriptive variables.

```{r , eval=TRUE}
colnames((mers_kor_14[[1]]))
```

Dataframe mapping relationship(s) between individuals (= network edges), unique identifiers and a set of descriptive variables.

```{r}
colnames((mers_kor_14[[2]]))
```

#### Create an `epi_contacts` object and get a summary

```{r}
# create epi_contacts object as directed network
x <- make_epi_contacts(mers_kor_14[[1]], mers_kor_14[[2]], directed=TRUE )

# summary of epi_contacts object
summary(x)
```

#### Interactive visualization  (generic methods)

The default plot methods uses the `visNetwork` package functionality to enable interactive visualization.

```{r, eval=FALSE}
plot(x)  # generic method, equivalent to the following: vis_epi_contacts(x)
```

#### Selection functions by individuals

Access unique IDs identifier(s) from `epi_contacts` object with various options:

1. **Based on list of case (node)**

```{r}  
# linelist only ('linelist')
linelist_ids <- get_id(x, "linelist")

# firt 10 records    
head(linelist_ids, n = 10)              

# calculate number of unieque 
length(linelist_ids)

# extract ids of contacts i
contacts_ids <- get_id(x, "contacts")
head(contacts_ids, n = 10)

# calculate number of records in contact list
length(get_id(x, "contacts"))  

# the union of both ('all')
all_ids <- get_id(x, "all") 
head(all_ids, n = 10)
length(get_id(x, "all"))

# Intersection of both ("common")
common_ids <- get_id(x, "common") 
head(common_ids, n = 10)
length(get_id(x, "common"))         
```

2. **Based on list of pairs (edge of the network)**
By default in a `epi_contacts` object, deg_in and deg_out returns the number of contacts/edges  per individuals/nodes. deg_in and deg_out requires a directed network.
[* ? to be tested with various direted and undirected network]

```{r}  

deg_in   <- get_degree(x, "in")

head(deg_in, n = 10)

# Distribution of incomning edge(s) by node
table(deg_in)           

deg_out   <- get_degree(x, "out")

head(deg_out, n = 10)

# Distribution of outgoing edge(s)by node
table(deg_out)

deg_both  <- get_degree(x, "both")

# ID record with highest number of edges
head(deg_both, n = 10)

# Distribution of edge numbers  by nodes
table(deg_both)
```

#### Extract Characteristics of Pairwise Nodes

The function allows pairwise processing of varaible in the line list according to each pairs in the contact dataset. For this example, date of onset of disease is extracted from the line list in order to compute the difference between disease date of onset date for each pairs (which corresponds to the **serial interval**).
[ ! add details of the default settings, behavior for date, string, numeric]

1. **By a date in the line list**

```{r} 
SI <- get_pairwise(x, "dt_onset")   
summary(SI)
```

Distribution of the serial interval:

```{r}
hist(SI, col="grey", border="white", xlab="Days after symptoms",
    main="MERS Korea 2014 - Serial Interval")
```

2. **By a categorical variable (sex in the example below) and applying a specfic function (e.g. table)**

```{r, eval=TRUE} 
get_pairwise(x, "sex", f=table) # allow the use of a custom function
fisher.test(get_pairwise(x, "sex", f=table)) 
```

<br>

#### Incorporate Cluster IDs and Sizes into the Line List

The function `get_clusters` adds a unique cluster identifier to the line list as a column named `cluster_member`. It also adds another column, `cluster_size`, corresponding to the size of the cluster. 
_Note: in this context,a  'cluster'  corresponds to groups of connected individuals (or connected components)._

```{r, message=FALSE}
# adds two columns, one with cluster IDs and other with cluster sizes
x_id <- get_clusters(x)

# one cluster of 94 observations connected together with at least one edge, one cluster of 3 observations  and 65 single observations.
table(x_id$linelist$cluster_size,x_id$linelist$cluster_member) 
```

##### Subsetting clusters in `epi_contacts` object by individual IDs or cluster size

1. **Subset network via node-based selection** 

```{r, eval=TRUE}
# select  one or more record(s)
nodes <- c("SK_14","SK_145")                  

# subset connected components including the proposed ID
x_subset_N <- subset_clusters_by_id(x, nodes)

# plot the 
plot(x_subset_N)                              
```
  
2. **Subset network via cluster size** 
  
```{r, eval=TRUE}   
  x_subset_S1 <- subset_clusters_by_size(x, cs = 3) # to be updated
  plot(x_subset_S1)
  x_subset_S2 <- subset_clusters_by_size(x, cs_min = 10, cs_max = 100)
  plot(x_subset_S2)
```

#### 2D Network Representation

```{r, eval=FALSE}
# Geoms to plot networks with ggplot2  
devtools::install_github("briatte/ggnetwork")
```

The current example is built is using `ggnetwork` package

```{r, eval=TRUE}  

library(igraph) 

# convert object for igraph processing
net_igraph <- as.igraph.epi_contacts(x)

# tree-like graph using the Reingold-Tilford graph layout algorithm
l <- layout.reingold.tilford(net_igraph)

# alternative circular layout
# l <- layout.reingold.tilford(net_igraph, circular=T)

plot(net_igraph,
    # NODES   
    vertex.color= "red" ,                                               
    vertex.frame.color= "white",                              
    vertex.shape= "circle",
    # size of the node (default is 15) or V(net)$size
    vertex.size = 2.5, 
    # character vector to name the nodes
    vertex.label = NA,                                      
    # EDGES
    # edge color "grey70"
    edge.color = E(net_igraph)$color,
    # edge width, defaults to 1
    edge.width = 0.5,
    # arrow size, defaults to 1
    edge.arrow.size = 0.1,
    # arrow width, defaults to 1
    edge.arrow.width = 0.25,    
    # empty space margins around the plot, vector with length 4
    margin = 0, 
    # if TRUE, the plot will be framed
    frame =  TRUE,
    # add plot title and format
    main = list(paste("R igraph package",sep=""), cex = 1, col = "black", font = 3),
    # specify layout chosen above
    layout=l) 
```
  
The current example is built is using `ggnetwork` package:

```{r}

library(ggplot2)
library(ggnetwork)
library(intergraph)

net_igraph  <-  as.igraph.epi_contacts(x_subset_N)

# For more information on conversion process
net_ggnet <- asNetwork(net_igraph)

# ggnetwork() function ensure conversion in dataframe for further used in ggplot2

# head(ggnetwork(net_ggnet))
# tail(ggnetwork(net_ggnet))
  
ggplot(ggnetwork(net_ggnet, layout = "fruchtermanreingold", arrow.gap = 0.01),
    aes(x, y, xend = xend, yend = yend)) +
    geom_edgetext(aes(label = diff_dt_onset ), label.size = NA, color = "grey10", fill = "white",size= 2.5, fontface = "bold") +
    geom_edges(aes(linetype = "solid"), size = 0.5, color= "grey60", alpha=0.7, arrow = arrow(angle= 40, length = unit(3, "pt"), type = "closed")) +
    geom_nodes(aes(), color = "goldenrod3", size=4, alpha=0.5) +
    theme_blank(legend.position="none")
```

#### 3D Network Representation

```{r}
library(threejs)

x_subset <- subset_clusters_by_size(x, cs_min = 50, cs_max = 100)
g <- graph3D(x_subset, v_col_by = "sex")

renderGraph(g)
```

```{r}
library(threejs)
x_subset <- subset_clusters_by_size(x, cs_min = 50, cs_max = 100)

graph3D(x_subset, v_col_by = "sex")
```

To interact with the plot:

* zoom: scrollwheel
* rotate: left-mouse button + move
* pan: right-mouse button + move
* identify node by label: mouse over 

For more information on 3D javascript graphs, visit B.W. Lewis' [page](http://bwlewis.github.io/rthreejs/graphjs.html)

#### `epicontacts` Shiny App

`epicontacts` includes an interactive mode that features a web interface built and designed using the [Shiny](http://shiny.rstudio.com/) framework. The "server" runs on a local host, and therefore can be viewed and used inside a web browser. With the package loaded, the app can be launched using the following:

```{r, eval=FALSE, message = FALSE}
epicontacts_server()
```
