---
title: "`epicontacts`: Manipulation, Visualisation and Analysis of Epidemiological Contact Data "
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
toc_depth: 4
vignette: >
    %\VignetteIndexEntry{Vignette Title}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r init, include=F}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE)
```

## Introduction   

`epicontacts` aims to facilitate manipulation, visualisation and analysis of epidemiological contact data. Such datasets inherently have network components (i.e. nodes and edges), and this package provides a convenient data structure as well as functionality specific to contact tracing and transmission chain analysis.

### Overview

At minimum, `epicontacts` requires input data partitioned as follows:

1. A **line list** of individuals containing records (i.e. nodes) of cases with unique identifiers of individuals and columns for additional attributes such as age, sex, location, etc.

2. A **contact** dataset that maps the relationships between individuals (i.e. edges between nodes), and can contain any attributes qualifying each relationship such as exposure, serial interval, etc.

**nb** for more on how to create an `epi_contacts` object see [loading data](#loading-data).

### Contents

The package includes the following features:

- **`epi_contacts`**: S3 class for storing line lists and contact datasets
- **`make_epi_contacts()`**: a constructor for the `epi_contacts` class
- **`get_id()`**: access unique IDs in an `epi_contacts` with various options
- **`get_pairwise()`**: extract attributes of record(s) in contacts database using information provided in the linelist data of an `epi_contacts` object.
- **`get_degree()`**: access degree of cases in `epi_contacts` with various options
- **`x[i,j,contacts]`**: subset an `epi_contacts` object by retaining specified cases
- **`subset()`**: subset an `epi_contacts` object by attributes or cluster characteristics
- **`summary()`**: summary for  `epi_contacts` objects
- **`plot()`**: plot for  `epi_contacts` objects; various types of plot are available; default to `vis_epi_contacts()`
- **`vis_epi_contacts()`**: plot an `epi_contacts` object using `visNetwork`
- **`as.igraph.epi_contacts()`**: create an `igraph` object from an epi_contacts object
- **`get_clusters`**: assign clusters and corresponding cluster sizes to linelist of an epi_contacts object
- **`graph3D()`**: 3D graph from an `epi_contacts` object
- **`epicontacts_server()`**: run the epicontacts Shiny app on a local host
- **`ebola.sim`**: simulated Ebola outbreak data in list format
- **`mers_kor_14`**: 2014 MERS outbreak data from South Korea in list format

## Installation

### GitHub

The latest development version of `epicontacts` is available on GitHub[^1], and can be installed using the `devtools` package.

```{r eval=FALSE} 
library(devtools)
install_github("Hackout3/epicontacts")
```

### CRAN

Alternatively, a stable release is availabe for installation via CRAN.

```{r, eval=FALSE}
install.packages("epicontacts")
```

Once installed, the package is ready for use.

```{r} 
library(epicontacts)
```

## Loading Data

`epicontacts` provides a convenient structure to store heterogeneous epidemiological contact network data (i.e. nodes and edges) in a single object. The `epi_contacts` class must contain two components: a line list and a contact dataset. 

Each row of the line list should represent unique observations of cases, and each row of the contact list should represent unique pairs of contacts. Each can include arbitrary features, but both datasets should share an identification scheme.

### Example Dataset

The example that follows will use the `mers_kor_14`, which is a dataset (in *list* format) that is included in the package. 

```{r}
str(mers_kor_14)
```

What features are in the line list?

```{r}
colnames((mers_kor_14[[1]]))
```

What about the contact dataset?

```{r}
colnames((mers_kor_14[[2]]))
```

### Creating `epi_contacts` Object

In order to create the `epi_contacts` object, both the line list and contact data frames must be passed to `make_epi_contacts()`. This function accommodates instances when the respective identifiers are not the first columns of these data frames (see the "id", "from" and "to" arguments). `make_epi_contacts()` can also account for contact networks that have a direction (see "directed" argument).

```{r}
merskor14 <- make_epi_contacts(linelist = mers_kor_14[[1]], contacts = mers_kor_14[[2]], directed = FALSE)
class(merskor14)
summary(merskor14)
```

## Data Manipulation

### Access Unique Identifiers

The `summary()` method above provided counts for the number unique cases in both the contact and line list. The `get_id()` function retrieves similar information but as vectors of identifiers. This can be parameterized as follows:

- **linelist**: IDs in the line list data
- **contacts**: IDs in the contact dataset ("from" and "to" combined)
- **from**: IDs in the "from" column of contact datset
- **to** IDs in the "to" column of contact dataset
- **all**: IDs that appear anywhere in either dataset
- **common**: IDs that appear in both contacts dataset and line list
    
What are the first ten IDs in the contacts dataset?
```{r}  
contacts_ids <- get_id(merskor14, "contacts")
head(contacts_ids, n = 10)
```

How many IDs are common to both?

```{r}
length(get_id(merskor14, "common"))
```

### Subsetting

The `subset()` method for `epi_contacts` objects allows for, among other things, pruning of networks based on values of node and edge attributes. These values must be passed as named lists to the respective argument.

```{r}
subset(merskor14, node.attribute = list("outcome" = "Dead", "sex" = "M"), edge.attribute = list("exposure" = "Emergency room"))
```

In addition to subsetting by node and edge attributes, networks can be pruned to only include components that are connected to certain nodes. The "id" argument takes a vector of nodes and returns the line list of individuals that "touch" those IDs.

```{r}
nodes <- c("SK_14","SK_145")                  
subset(merskor14, cluster_id = nodes)
```

The `subset()` method for `epi_contacts` objects also accepts cluster size parameters (see "cs", "cs_min" and "cs_max" arguments).
    
```{r}   
subset(merskor14, cs = 3)
subset(merskor14, cs_min = 10, cs_max = 100)
```

## Visualisation

### Default Plotting Method

One of the main features of `epicontacts` is its visualisation capabilities. As a default, the package uses interactive plotting based on the `visNetwork` package[^2]. This interactivity is particularly useful for visualising large datasets.

```{r}
plot(merskor14) 
```

The above is a generic method based on the `vis_epi_contacts()` and accepts a number of arguments to customize the plot appearance and functionality. For a full list of options use `?vis_epi_contacts()`

### 3D

`epicontacts` loads the `threejs` package to enable 3D visualisation tools with the `graph3D()` function[^3].

```{r}
graph3D(merskor14, v_col_by = "sex", v.size = 0.5, g.title = "MERS Korea 2014")
```

To interact with the plot:

- **zoom**: scrollwheel
- **rotate**: left-mouse button + move
- **pan**: right-mouse button + move
- **identify node by label**: mouse over 

### Shiny Interface

`epicontacts` includes an interactive mode that features a web interface built and designed using the Shiny framework[^4]. The "server" runs on a local host, and therefore can be viewed and used inside a web browser. With the package loaded, the app can be launched with the `epicontacts_server()` function.

```{r, eval=FALSE}
epicontacts_server()
```

## Analysis

### Extract Characteristics of Pairwise Nodes

The `get_pairwise()` function allows processing of variable(s) in the line list according to each pair in the contact dataset. For the following example, date of onset of disease is extracted from the line list in order to compute the difference between disease date of onset for each pair. The value that is produced from this comparison represents the **serial interval (si)**.

```{r} 
si <- get_pairwise(merskor14, "dt_onset")   
summary(si)
hist(si, col="grey", border="white", xlab="Days after symptoms",
    main="MERS Korea 2014\nSerial Interval")
```

The `get_pairwise()` will interpret the class of the column being used for comparison, and will adjust its method of comparing the values accordingly. For numbers and dates (like the **si** example above), the function will subtract the values. When applied to columns that are characters or categorical, `get_pairwise()` will paste values together. Because the function also allows for arbitrary processing (see "f" argument), these discrete combinations can be easily tabulated and analyzed.
    
```{r} 
head(get_pairwise(merskor14, "sex"), n = 10)
get_pairwise(merskor14, "sex", f=table)
fisher.test(get_pairwise(merskor14, "sex", f=table)) 
```

### Identify Clusters

The `get_clusters()` function returns an augmented `epi_contacts` object that includes two extra features in the line list: *cluster_member* and *cluster_size*. In this context, a 'cluster' corresponds to groups of connected individuals or components.

```{r}
names(get_clusters(merskor14)$linelist)
```

### Calculate Degree

The degree of a node corresponds to its number of edges or connections to other nodes. `get_degree()` provides an easy method for calculating this value for `epi_contacts` networks. A high degree in this context indicates an individual who was in contact with many others.

**nb** use of "type" argument depends on whether or not the network is directed. 

```{r}  
deg_both <- get_degree(merskor14, "both", only_linelist = TRUE)
```

Which individuals have the ten most contacts?

```{r}
head(sort(deg_both, decreasing = TRUE), 10)
```

What is the mean number of contacts?

```{r}
mean(deg_both)
```

## References

[^1]: https://github.com/reconhub/epicontacts
[^2]: https://cran.r-project.org/web/packages/visNetwork/vignettes/Introduction-to-visNetwork.html
[^3]: http://bwlewis.github.io/rthreejs/graphjs.html
[^4]: http://shiny.rstudio.com/