---
title: "`epicontacts` Class: Details Regarding the Data Structure for `epicontacts` Objects"
author: "VP Nagraj"
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
     toc: true
     toc_depth: 4
vignette: >
  %\VignetteIndexEntry{epicontacts class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, include=F}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE)
```

## `make_epicontacts()`

The `epicontacts` data structure is useful for epidemiological network analysis of cases and contacts. Data partitioned as *line list* and *contact list* formats can be coerced to the `epicontacts` class in order to facilitate manipulation, visualization and analysis. 

The `make_epicontacts()` function creates this data structure, and accepts arguments for:

- **linelist**: a data frame with at least one column providing unique case identifiers
- **contacts**: a data frame with at least two columns indicating individuals that were in contact with one another (*nb* these may or may not be referenced in the line list)
- **id**: the name or index for the column in the linelist that contains unique identifiers for individual cases; defaults to the first column
- **from**: the name or index for the column in the contacts that contains the originating case; defaults to first column; can still be used in non-directed networks but should not be interpreted as direction (see **directed** argument)
- **to**: the name or index for the column in the contacts that contains secondary case; defaults to second column; can still be used in non-directed networks but should not be interpreted as direction (see **directed** argument)
- **directed**: indicator as to whether or not the contact moved in a direction (i.e. "from" one individual "to" another)

Using a simulated ebola outbreak dataset rom the **outbreaks** package, we can explore how to create an `epicontacts` object.

```{r}
library(outbreaks)
library(epicontacts)

str(ebola_sim$linelist)
str(ebola_sim$contacts)

x <- make_epicontacts(linelist = ebola_sim$linelist, contacts = ebola_sim$contacts, directed = TRUE)
```

```{r}
is.list(x)
class(x)
```


## `$linelist`

```{r}
x$linelist
```


## `$contacts`

```{r}
x$linelist
```


## Generics

### `print.epicontacts()`

```{r}
x
```

### `summary.epicontacts()`

```{r}
summary(x)
```

### `subset.epicontacts()`

```{r}
subset(x, node.attribute = list(""))
```

### `plot.epicontacts()`


