---
title: "Epicontacts: visualisation of network data in epidemiology"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
runtime: shiny
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


<br>
Epicontacts package intends to facilitate the visualisation and analysis of any epdimiological data with an network component, such as contact tracing datasets and  transmission chains.

Inputs data is built upon:

<bk>

 1. a **line list** (LL) of individuals.The line list contains a first column with unique identifier of individuals (record of cases and/or contacts) and any attributes in additionnal columns (e.g. age, sex, location, ...).

<bk>

2. a **contact dataset** (CD) mapping the relationships between individuals. The contact dataset corresponds to the mapping of the relationships between individuals (edge between nodes). The contact dataset can contains any attributes with regards to this relationship (e.g.  exposure, serial interval, ...).

<bk>

The package provides several network statistics,  functions to subset networks (by edge or two node) and visualize network in a various ways (2D, 3D, interactive graphics) using third-parties packages .

---------------------------------------
  
### Installation
  
To install the development version from github:
```{r install, eval=FALSE} 
  library("devtools")
  install_github("Hackout3/epicontacts")
```
  
The stable version can be installed from CRAN using: [*to be checked if needed*] _{Nistara: package is not on CRAN yet}_
```{r, eval=FALSE}
  # install.packages("epicontacts",force = TRUE) # ! Error: attempt to use zero-length variable name 
```

Then, to load the package, use:
```{r, eval=TRUE} 
  library("epicontacts")
```
  
### Content overview

Main features include:

* **epi_contacts**: a new S3 class for storing line list and contacts datasets.

* **as.igraph.epi_contacts**: creates an `igraph` object from a `epi_contacts` object.

* **clusters_epi_contacts**: assign clusters IDs to line list of an `epi_contacts` object (clusters being groups of connected individuals/nodes).

* **ebola.sim**: simulated Ebola outbreak dataset. 

* **get_degree**: number of contacts/edges  per individuals/nodes in a `epi_contacts` object.

* **get_id**:access unique IDs identifier(s) from `epi_contacts` object with various options.

* **get_pairwise**:  extract attributes of record(s) in contacts database using information provided in the line list database of an `epi_contacts` object.

* **graph3D**: 3D graph from an `epi_contacts` object.

* **make_epi_contacts**: a constructor for the new `epi_contacts` class.

* **subset_clusters_by_nodes**: subset an `epi_contacts` object based on individuals/nodes of interest.

* **subset_clusters_by_size**:  subset an `epi_contacts` object based on size(s) of clusters (clusters being groups of connected individuals/nodes).

* **vis_epi_contacts**: plots epi_contacts objects using the visNetwork package.

  
Other generic functions are proposed:

* **summary**: summary for  `epi_contacts` object.

* **plot**:  default plot for `epi_contacts` object using visNetwork package.

* **print**: default printing method for `epi_contacts` object.

* **x[i,j,contacts]**: susbset an `epi_contacts` object by retaining specified cases. [*to be checked if needed*]

<br>

Overview of package contents:

```{r Overv, eval=TRUE}
ls("package:epicontacts")
```

<br>
Load additionnal packages required for running examples below: 
loaded igraph, visNetwork

```{r lib, eval=TRUE}
  # library("intergraph")
  library("shiny")
  library("ggplot2")
```

<br>

### Visualisation and analysis

##### Load training dataset
  
A new S3 class `epi_contacts` for storing line list (LL)  and contacts dataset (CD) can be created from a line list (one row = one individual) and a contact dataset containing the relationship betwen individuals (one row = one edge). A training dataset `mers_kor_14` is used as running example (for more information, see  ` outbreaks` package  for dataset description).

```{r , eval=TRUE}
  str(mers_kor_14) # list of two dataframes (one line list and one contact dataset)
```

Dataframe corresponding to a line listing of individuals (= network node) which can be case(s) and/or contact(s) (if any), unique identifier and a set of descriptive variables.

```{r , eval=TRUE}
  str(mers_kor_14[[1]])
  colnames((mers_kor_14[[1]]))
  summary(mers_kor_14[[1]])
```

Dataframe mapping relationship(s) between individuals (= network edge), unique identifiers and a set of descriptive variables.

```{r , eval=TRUE}
  str(mers_kor_14[[2]]) 
  colnames((mers_kor_14[[2]]))
  summary(mers_kor_14[[2]])  # chekc function  NULL:SK_14 in from and  NULL:SK_113 in to
```

<br>

### Create an epi_contacts object and get a summary

```{r, eval=TRUE}
  x <- make_epi_contacts(mers_kor_14[[1]],mers_kor_14[[2]], 
                         id="id", to="to", from="from",  
                         directed=TRUE )    # create epi_contacts object as directed network
  
  summary(x)        # summary of epi_contacts object
```

<br>

#### Interactive visualization  (generic methods)

The default plot methods is using the "visNetwork" package to allow interactive visualization of network.

```{r, eval=FALSE}
  plot(x)  # generic method, equivalent to the following: vis_epi_contacts(x)
```

<br>


#### Selection functions by individuals

Access unique IDs identifier(s) from epi_contacts object with various options:
  
* Based on list of case (node)

```{r, eval=TRUE}  
linelist_ids <- get_id(x, "linelist")   # linelist only ('linelist')
head(linelist_ids, n = 10L)
length(linelist_ids)                    # 162 unique records in line list

contacts_ids <- get_id(x, "contacts")   # contacts only ('contacts')
head(contacts_ids, n = 10L)
length(get_id(x, "contacts"))           # 97 unique records in line list

all_ids <- get_id(x, "all")             # the union of both ('all')
head(all_ids, n = 10L)
length(get_id(x, "all"))

common_ids <- get_id(x, "common")       # Intersection of both ('common')
head(common_ids, n = 10L)
length(get_id(x, "common"))         
```
  
* Based on list of pairs (edge of the network)
By default in a `epi_contacts` object, deg_in and deg_out returns the number of contacts/edges  per individuals/nodes. deg_in and deg_out requires a directed network.
[* ? to be tested with various direted and undirected network]

```{r, eval=TRUE}  
  deg_in   <- get_degree(x, "in")
  head(deg_in, n = 10L)      
  table(deg_in)             # Distribution of incomning edge(s) by node
  
  deg_out   <- get_degree(x, "out")
  head(deg_out, n = 10L)     
  table(deg_out)            # Distribution of outgoing edge(s)by node
  
  deg_both  <- get_degree(x, "both")
  head(deg_both, n = 10L)    #  ID record with highest number of edges
  table(deg_both)           # Distribution of edge numbers  by nodes

```

<br>

#### Extract characteristics of pairwise nodes

The function allows pairwise processing of varaible in the line list according to each pairs in the contact dataset. For this example, the onset of disease is extracted from the line list in order to compute the difference between disease onset date for each pairs (which corresponds to the **serial interval**).
[ ! add details of the default settings, behavior for date, string, numeric]

* By a date in the line list

```{r, eval=TRUE} 
SI <- get_pairwise(x, "dt_onset")   
summary(SI)

```
<br>

Distribution of the serial interval:

```{r, eval = TRUE}
renderPlot({
    hist(SI, col="grey", border="white", xlab="Days after symptoms",
     main="MERS Korea 2014 - Serial Interval")
    })  # no dysplay to 
```

* By a categorical variable (sex in the example below) and applying a specfic function (e.g. table)

```{r, eval=TRUE} 
get_pairwise(x, "sex", f=table) # allow the use of a custom function
fisher.test(get_pairwise(x, "sex", f=table)) 
```

<br>
####  Incorporate cluster IDs and sizes into the line list
The function `clusters_epi_contacts` adds a unique cluster identifier to the line list as a column named `cluster_member`. It also adds another column, `cluster_size`, corresponding to the size of the cluster. 
_Note: A 'cluster' in this context corresponds to groups of connected individuals (connected components)._

```{r, eval=TRUE, message=FALSE} 
x_id <- clusters_epi_contacts(x)    # adds two columns, one with cluster IDs and other with cluster sizes
head(x_id$linelist, n = 10)
head(x_id$linelist$cluster_member, n = 10)
head(x_id$linelist$cluster_size, n = 10)
table(x_id$linelist$cluster_member)         # table of clusters with the number of individuals in them
sort(unique(x_id$linelist$cluster_size))    # unique cluster sizes of clusters in linelist
```

<br>
##### Subsetting clusters in `epi_contacts` object by individual IDs, cluster member number or cluster size

* ** Subset network via  node-based selection** 

```{r, eval=TRUE}   
  nodes <- c("SK_14","SK_145")              # Select  one or more record(s)
  x_subset_N <- subset_clusters_by_nodes(x, nodes)  # connected component which includes the selected nodes
  plot(x_subset_N)            # selection  as cluster to which "SK_14" belongs = all ?
```
  
* ** Subset network via cluster size** 
  
```{r, eval=TRUE}   
  x_subset_S <- x_subset <- subset_clusters_by_size(x, cs_min = 2, cs_max = 3)
  plot(x_subset_S)
```

<br>

#### Network representation with static 2D representation 


```{r, eval=FALSE}
  install.packages("ggnetwork") ######### Do we need to install this if we're using github below?
  devtools::install_github("briatte/ggnetwork") # Geoms to plot networks with ggplot2  
```


The current example is built is using ggnetwork package

```{r, eval=TRUE}  
  
  library("igraph") 
  library("intergraph")
  library("ggnetwork")

  net_igraph  <-  as.igraph.epi_contacts(x)   # Conversion in i-graph object
  l <- layout.reingold.tilford(net_igraph)    # Placement algorithm: Reingold Tilford
  
  renderPlot({
      plot(net_igraph,layout=l, edge.arrow.size=.2, 
           vertex.label=NA,
           vertex.frame.color="#ffffff", 
           vertex.size=2)
  })   # Default plot from igraph --> ?? adapt layour
  
  net_ggnet <- asNetwork(net_igraph)
  # For more information on conversion process
  
  net_ggnet <- ggnetwork(net_ggnet)
  
  # Default plot from ggnet2
```
  
The current example is built is using ggnetwork package

```{r, eval=FALSE}
  
    ggplot(
      ggnetwork(net_igraph, layout = "kamadakawai", arrow.gap = 0.025),
      aes(x, y, xend = xend, yend = yend)) +
      geom_edges(aes(color = ser_int), arrow = arrow(length = unit(10, "pt"), type = "closed")) + # edge layout
      geom_nodes(aes(size = 3)) + # node layout
      scale_color_gradient(low = "grey50", high = "tomato") +
      scale_size_area(breaks = 1:3) +
      theme_blank()
    )
```

<br>
 
#### Network representation with static 3D representation

```{r, eval=TRUE}
library(threejs)

data(LeMis)
renderGraph( graphjs(LeMis))
```
test space

```{r, eval=TRUE}

z <- seq(-10, 10, 0.01)
x <- cos(z)
y <- sin(z)
scatterplot3js(x,y,z, color=rainbow(length(z)))

```

<br>

#### Run contacts package as a app.


<br>
 
#### Run contacts package as a app. ContactsServer: a web application 
  
  The functionalities of Contacts are also available via a user-friendly web interface,    running locally on the default web browser.   It can be started by simply typing ContactsServer(). The interface allows you to import both dataset required  epicontact object  (line list  and contacts data) and explore interactively network.
  --> Fonction export to be developped   ? saving image of current view in various formats. 
  


  [* ? To be completed]
The `epi_contacts` object can be subsetted in different manner using specific conditions:

```{r, eval=FALSE}  
  x[i,j,contacts]
```

